
\begin{thebibliography}{99}
\bibitem{beltrami}
E.~Beltrami,
\emph{Mathematical Models in the Social and Biological Sciences},
Jones and Bartlett Publishers, Boston, 1993.
%
\bibitem{Bender}
E.~A.~Bender,
\emph{An Introduction to Mathematical Modeling},
Dover Publications, Mineola, New York, 2000.
%
\bibitem{Braun}
M.~Braun, \emph{Differential Equations and Their Applications,
3rd Edition},
Springer-Verlag, New York, 1983.
%
\bibitem{GM}
J.~M.~Gottman, J.~D.~Murray, C.~C.~Swanson,
R.~Tyson, and K.~R.~Swanson,
\emph{The Mathematics of Marriage},
MIT Press, Cambridge, 2002.
%
\bibitem{Lanchester}
F.~W.~Lanchester,
\emph{Aircraft in warfare}, Appleton, New York, 1916.
%
\bibitem{Leslie45}
P.~H.~Leslie,
On the use of matrices in certain population mathematics,
\emph{Biometrika}, Vol. 33, No. 3, 183--212 (Nov., 1945)
%
\bibitem{Leslie48}
P.~H.~Leslie,
Some further notes on the use of matrices in population
mathematics,
\emph{Biometrika}, Vol. 35, No. 3/4, 213--245 (Dec., 1948)
%
\bibitem{LinSegel}
C.~C.~Lin and L.~A.~Segel,
\emph{Mathematics Applied to Deterministic
Problems in the Natural Sciences},
Classics in Applied Mathematics 1,
SIAM, Philadelphia, 1988.
%
\bibitem{NowackEvolutionaryDynamics}
M.~A.~Nowack,
\emph{Evolutionary Dynamics},
Harvard University Press,
Cambridge, Mass., 2006.
%
\bibitem{Olinick}
M.~Olinick,
\emph{An Introduction to Mathematical Models
in the Social and Life Sciences},
Addison-Wesley, Reading, 1978.
%
\bibitem{Rinaldi}
S.~Rinaldi, Laura and Petrarch: an intriguing case of
cyclical love dynamics,
\emph{SIAM J. Appl. Math.}, \textbf{58}(4), 1205--1221 (1998)
%
\bibitem{SCILAB}
\texttt{www.scilab.org}; Scilab is a registered trademark
of INRIA.
%
\bibitem{Solow}
R.~M.~Solow, A Contribution to the Theory of Economic Growth,
\emph{Quarterly Journal of Economics}, February, 1956, pp. 65--94.
\bibitem{Sprott}
J.~C.~Sprott,
Dynamical models of love,
\emph{Nonlinear Dynamics, Psychology, and Life Sciences},
\textbf{8}(3) (2004)
%
\bibitem{StrogatzRJ}
S.~H.~Strogatz, Love affairs and differential equations,
\emph{Mathematics Magazine}, \textbf{61}(1) (1988)
\end{thebibliography}
